Co-expression Network Layout
============================
2013/09/30

Overview
--------

Setup
-----
First, let's convert the edge list that has been formatted for use with
Cytoscape into a format supported by igraph.

```{r knitr_opts}
# Knitr defaults
library(knitr)
opts_chunk$set(fig.width=12, fig.height=12, fig.dpi=96)
```

```{r load_graph}
require(igraph)

# Load graph
g = read.graph('network.graphml', format='graphml')
head(V(g)$name)
head(E(g))

# To speed things up for now, let's limit ourselves to 100 vertices
num_keep = 2000
g = g - sample(V(g), length(V(g)) - num_keep)

# Let's remove any isolated vertices
g = g - names(degree(g)[degree(g) == 0])
```

First let's try plotting the network using the default layout
```{r network_plot_default}
plot(g, vertex.size=2, vertex.label=NA)
```

Next, let's try plotting the same network but using force-directed layouts.

```{r network_plot_force}
# Fruchterman-Reingold
coords = layout.fruchterman.reingold(g, weights=E(g)$weight)
plot(g, layout=coords, vertex.size=2, vertex.label=NA)

# Kamada-Kawai
coords2 = layout.kamada.kawai(g, weights=E(g)$weight)
plot(g, layout=coords2, vertex.size=2, vertex.label=NA)
```

Next, let's see what the effect of adjusting the edge weights is. Before doing
that, however, let's first look at the distribution of weights.

```{r edge_weights_orig, fig.height=9}
hist(E(g)$weight, breaks=15)
```

From the above histogram, it becomes apparent that the weights are all clustered
together around 0.6 - 0.65. This is due to the thresholding that was performed
on the original network data to reduce the number of edges before exporting:
in that case only edges >= 0.6 were kept.

First, let's try simply spreading the weights out over the range (0,1).

```{r edge_weights_scaled_1, fig.height=9}
w = E(g)$weight

scaled_weights = (w - min(w)) / (max(w) - min(w))
hist(scaled_weights, breaks=15)
```

Let's try using the new scheme on the same force-directed layout method
above.

```{r force_directed_scaled}
# Fruchterman-Reingold
coords3 = layout.fruchterman.reingold(g, weights=scaled_weights)
plot(g, layout=coords3, vertex.size=2, vertex.label=NA)

coords4 = layout.fruchterman.reingold(g, weights=scaled_weights * 100)
plot(g, layout=coords4, vertex.size=2, vertex.label=NA)

# Kamada-Kawai
coords5 = layout.kamada.kawai(g, weights=scaled_weights)
plot(g, layout=coords5, vertex.size=2, vertex.label=NA)

coords6 = layout.kamada.kawai(g, weights=scaled_weights * 100)
plot(g, layout=coords6, vertex.size=2, edge.width=0.25, vertex.label=NA)
```

